### 2018-03-23 paper/sexSelectionSpacing-ver1.tex
Full draft of conclusion

The main question addressed here is the extent to which spacing patterns 
significantly changed as prenatal sex determination became available,
specifically, did the spacing after girls increase relative to the spacing
after boys?
The underlying idea is that in the absence of sex selection son 
preference leads to shorter spacing after the birth of a girl than after the 
birth of a boy, whereas son preference increases the spacing after the birth 
of a girl relative to after the birth of a boy when prenatal sex selection is 
available.
The four rounds of India's NFHS provide strong support for the hypothesis.

Censoring is an important issue when trying to understand the development
in sex ratio, especially when it comes to establishing whether the
use of sex selection is increasing or decreasing.

This points to two important questions for future research:
What is the role of falling fertility in the use of sex selection
and is the use of sex selection increasing or decreasing over time?
[more on how to analyze those]

The results here lead to a host of larger important questions that
needs to be addressed.

First, to what extent are the improvements in health for girls
relative to boys the result of selection, longer spacing between
births, or changing son preference?

Second,

This suggestion that an overlooked approach to addressing a biased
sex ratio may be to encourage higher female labor force participation.
In addition to the obvious benefit to the women in terms of 
bargaining power, it also increases the cost of repeated 
sex-selective abortions.
[Caveat, it also lowers fertility, and everything else equal that
would may lead to higher use of sex selection]

>>>>

The main question addressed here is the extent to which spacing patterns 
significantly changed as prenatal sex determination became available.
Specifically, did the spacing after girls increase relative to the spacing
after boys?
The underlying idea is that in the absence of sex selection son 
preference leads to shorter spacing after the birth of a girl than after the 
birth of a boy, whereas son preference increases the spacing after the birth 
of a girl relative to after the birth of a boy when prenatal sex selection is 
available.
I introduce a new method that simultaneously account for spacing between
births and the potential use of sex selection and apply the method
to over four decades of data from India's NFHS.
The results show two very different approaches to son preference.

At one extreme, women without education mostly follow the standard 
pattern of shorter spacing when a woman does not have the desired number of 
son and with limited evidence of biased sex ratios and the increases in
spacing that should follow from the use of sex selection.
The situations where there are elevated sex ratios for women without
education---which, by itself, should indicate use of sex selection---the
spacing patterns are too close to the ones without sex selection that 
it is difficult to conclude that the changes in sex ratios indicate
sex selection rather than normal variation or problems with recall errors.

At the other extreme, women with eight or more years of education have
experienced an almost complete reversal of their spacing patterns.
Rather than having the next birth sooner as they mostly did before 
sex selection became available, women with either no or one son now 
have substantially longer spacing than if they have two or more sons.
That these changes are likely due to increased use of sex selection 
rather than changes in son preference as shown by the increasingly
significant deviations in the sex ratio from the natural sex ratio.

Over the last 40 years there has been a general increase in median
spacing for all education groups, but especially for the most 
educated women.
The exception is for the spacing between marriage and first birth
where there are two counteracting forces: women have become more likely 
to conceive after marriage and there has been a substantial compression 
in the variation.
The end result is that median spacing from marriage to first birth
that has changed little or even declined for the best educated women.

The empirical model is designed to allow for the effects of covariates
to change within the spell.
This is important because the use of sex selection is likely to vary
within a spell, either because preferences for spacing between births
that is not too long override the preference for sons or because
of selection where the women who remain in a spell are those with
the strongest son preference who have been through multiple 
pregnancies that were aborted because the fetus was female.
Because these move in opposite directions it is difficult to 
establish the extent to which they affect the results.
Both are, however, clearly present as can be seen from the
cases where there is either a decline in sex ratio with spell
duration or a substantial increase with spell duration.
The first case is more likely at lower parities where there
is still less pressure to ensure a son, which allows any
preference for shorter spacing to play more of a role.
The second is on the other hand more likely to have an effect
for higher parities where the desire to limit further
childbearing and ensure a son dominates any preferences for
shorter spacing.

The potential for changes in sex ratio within a spell means
that censoring is an important issue when trying to understand the development
in sex ratio.
This is especially the case when it comes to establishing whether 
the use of sex selection is increasing or decreasing.
Take, for example, the case of women with eight or more years of
education in the last period. 
If we look at women who started their third spell in either the
2005--2009 or the 2010--2014 periods and we treat end of 2015
as the cut-off time, then the observed sex ratio for the 2005--2009
women is 55.99\% boys while it is 55.38\% for women who started 
in the 2010--2014 period.%
\footnote{
This combines urban and rural women; the basis result 
changes little if I split by area of residence.
I chose the 2014 cut-off to ensure that both groups cover 
five years, even though NFHS-4 took place in 2015 and 2016.
}
It would therefore appear that the use of sex selection for
this spell is slightly decreasing over time.
This interpretation, however, ignores the role of censoring.
If we use the estimation results to predict final sex ratio
once childbearing for the third spell is over, the predicted
sex ratios are 56.10\% boys and 56.08\% boys for the 
2005--2009 and the 2010--2015 starts.
The reason for the difference is that while for the 
2005--2009 start we would have observed 7,134 birth out
of a predicted 7,451 total number of births, we would
only have observed 3,348 births out of a predicted 8,842
for the 2010--2014 group.

Although it is not possible to directly observe whether the
unequal sex ratio and the longer spacing patterns are caused
by sex selection the results are consistent with a general
increase in sex selection over time.
The increase in apparent use of sex selection is mainly found
among women with more education and more so in urban than in
rural areas, although the differences between urban and rural
areas are relatively minor.
Compared to previous research one result stands out.
It appears that the use of sex selection is spreading to 
women with lower levels of education, so that even women
with one to seven years of education show unequal sex ratios
and longer spacing for their higher-order spells.

This points to two important, interconnected, questions for 
future research:
What is the connection between falling fertility and the 
use of sex selection and is the use of sex selection increasing 
or decreasing over time?
It is clear that fertility has been declining over time in India,
although it is important to note that the decline might be
overstated exactly because of an increased use of sex selection and
the resulting longer spacing between births makes it less likely 
that we will observe births to women, even though they eventually
will happen.
The facts that biased sex ratio occur at earlier spells over
time and starting to show up for women with less education 
all indicate that lower fertility and use of sex selection are
related, but a more detailed analysis is needed.
Furthermore, although examining spells separately, as done here, is a useful
first step in understanding how sex selection changes over time,
it does not lend itself easily to examine the degree to which
sex selection changes over time.
For that a combined model that incorporates the full fertility
history of women would be more appropriate.
A model of this type would allow us to jointly model fertility
and sex selection and help address both questions.

The results here also lead to a set of broader important questions that
need to be addressed in future research.
First, to what extent are the improvements in health for girls
relative to boys the result of selection, longer spacing between
births, or changing son preference?
There is some early evidence that selection make it appear that
girls are healthier, even though the real cause is a combination
of sex selection and higher mortality together with recall error \citep{Portner2018a}. 
Understanding what role these factors play in better health for
girls is especially important when evaluating policies that 
aim to directly limit the use of sex selection rather than 
changing preferences or incentives.

Second, we need to understand how female labor force participation
interact with the use of sex selection.
Increased autonomy for women, through, for example, better opportunities
for working outside the home, has been suggested as a way to increase
women's status and thereby lowering the use of sex selection \citep{Das-Gupta2016}.
This may, however, be a double-edged sword.
On the one hand, it is an obvious benefit to the women in terms of 
bargaining power and it also increases the cost of repeated 
sex-selective abortions.
On the other hand, it may also further lower fertility, and everything 
else equal, that may lead to higher use of sex selection.

[Nothing is ever simple!]

This suggestion that an overlooked approach to addressing a biased
sex ratio may be to encourage higher female labor force participation.
